---
title: "project PCA control variables"
author: "Nick Kachanyuk"
date: "7/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

# Removing non-control variables

```{r}
PCA_variables <- train %>% select(-amount_tsh, -public_meeting, -permit, -management_group, -payment_type, -literacy_rate_by_region, -pct_unemployed_by_region, - well_strain)

# these will end up being the non-control variables
```

# PCA

```{r}
pr_train <- prcomp(x = select(PCA_variables,-status_group), scale = T, center = T)
summary(pr_train)
```

```{r}
screeplot(pr_train, type = "lines")
```

```{r}
pr_train$rotation
```

```{r}
library(ggbiplot)
```

```{r}
ggbiplot(pr_train, ellipse = TRUE, groups=train$status_group) + coord_equal(ratio=0.25)
```

# PC 1

```{r}
prc <- bind_cols(select(train,status_group),as.data.frame(pr_train$x))

prc %>% select(status_group, PC1) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 2

```{r}
prc %>% select(status_group, PC2) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 3

```{r}
prc %>% select(status_group, PC3) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 4

```{r}
prc %>% select(status_group, PC4) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

The graphs above show that the 4 principal components are not really great at separating the 3 functional groups from each other. This is also evident when looking at how much variance each PC captures where the PC 1 with the highest variance is about 11 %. 

I try to fix this issue by further subsetting the PCR_variables into the following categories: geographic characteristics, demographics, characteristics of well

# Geographic PCA

```{r}
geographic_PCA <- PCA_variables %>% select(status_group, gps_height, longitude, latitude, region_area_sq_mi, basin_Internal, `basin_Lake Nyasa`, `basin_Lake Rukwa`, `basin_Lake Tanganyika`, `basin_Lake Victoria`, basin_Pangani, basin_Rufiji, `basin_Ruvuma / Southern Coast`, `basin_Wami / Ruvu`, region_Arusha, `region_Dar es Salaam`, region_Dodoma, region_Iringa, region_Kagera, region_Kigoma, region_Kilimanjaro, region_Lindi, region_Manyara, region_Mara, region_Mbeya, region_Morogoro, region_Mtwara, region_Mwanza, region_Pwani, region_Rukwa, region_Ruvuma, region_Shinyanga, region_Singida, region_Tabora, region_Tanga)
```

```{r}
pr_geo <- prcomp(x = select(geographic_PCA,-status_group), scale = T, center = T)
summary(pr_geo)
```

```{r}
screeplot(pr_geo, type = "lines")
```

```{r}
pr_geo$rotation
```

```{r}
ggbiplot(pr_geo, ellipse = TRUE, groups=train$status_group) + coord_equal(ratio=.85)
```

```{r}
pca_geo_df <- rownames_to_column(as.data.frame(pr_geo$rotation)) %>%
  select(1:5) %>%
  filter(abs(PC1) >= 0.25 | abs(PC2) >= 0.25 | abs(PC3) >= 0.25 | abs(PC4) >= 0.25)

pca_geo_df
```

```{r}
save(pca_geo_df,file="pca_geo_df.Rda")
#load("pca_geo_df.Rda")
```

```{r}
# pc1_geo is south eastern Tanzanian regions that are moderate to large region area sq mi
# pc2_geo is north western Tanzania region that are small to moderate region area sq mi
```

# PC 1

```{r}
prc_geo <- bind_cols(select(train,status_group),as.data.frame(pr_geo$x))

prc_geo %>% select(status_group, PC1) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 2

```{r}
prc_geo %>% select(status_group, PC2) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 3

```{r}
prc_geo %>% select(status_group, PC3) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 4

```{r}
prc_geo %>% select(status_group, PC4) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

Again there is much separation of the status groups by geographical principal components.

# Demographic PCA

```{r}
demographic_PCA <- PCA_variables %>% select(status_group, population, total_region_pop, households_per_region, mhead_avg_household_size_per_region, fhead_avg_household_size_per_region, region_pop_density)
```

```{r}
pr_demo <- prcomp(x = select(demographic_PCA,-status_group), scale = T, center = T)
summary(pr_demo)
```

```{r}
screeplot(pr_demo, type = "lines")
```

```{r}
pr_demo$rotation
```

```{r}
ggbiplot(pr_demo, ellipse = TRUE, groups=train$status_group) + coord_equal(ratio=.85)
```

```{r}
pca_demo_df <- rownames_to_column(as.data.frame(pr_demo$rotation)) %>%
  select(1:5) %>%
  filter(abs(PC1) >= 0.25 | abs(PC2) >= 0.25 | abs(PC3) >= 0.25 | abs(PC4) >= 0.25)

pca_demo_df
```

```{r}
save(pca_demo_df,file="pca_demo_df.Rda")
#load("pca_demo_df.Rda")
```

```{r}
# pc1_demo is highly populated wells, small region population and density
# pc2_demo is less populated wells, slightly more populated region
```

# PC 1

```{r}
prc_demo <- bind_cols(select(train,status_group),as.data.frame(pr_demo$x))

prc_demo %>% select(status_group, PC1) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 2

```{r}
prc_demo %>% select(status_group, PC2) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 3

```{r}
prc_demo %>% select(status_group, PC3) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 4

```{r}
prc_demo %>% select(status_group, PC4) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 5

```{r}
prc_demo %>% select(status_group, PC5) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 6

```{r}
prc_demo %>% select(status_group, PC6) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# Well characteristic PCA

```{r}
characteristics_PCA <- PCA_variables %>% select(status_group, construction_year, extraction_type_class_gravity, extraction_type_class_handpump, extraction_type_class_motorpump, extraction_type_class_other, `extraction_type_class_rope pump`, extraction_type_class_submersible, `extraction_type_class_wind-powered`, quality_group_colored, quality_group_fluoride, quality_group_good, quality_group_milky, quality_group_salty, quality_group_unknown, quantity_group_dry, quantity_group_enough, quantity_group_insufficient, quantity_group_seasonal, quantity_group_unknown, source_type_borehole, source_type_dam, source_type_other, `source_type_rainwater harvesting`, `source_type_river/lake`, `source_type_shallow well`, source_type_spring, source_class_groundwater, source_class_surface, source_class_unknown, `waterpoint_type_group_cattle trough`, `waterpoint_type_group_communal standpipe`, waterpoint_type_group_dam, `waterpoint_type_group_hand pump`, `waterpoint_type_group_improved spring`, waterpoint_type_group_other)
```

```{r}
pr_chara <- prcomp(x = select(characteristics_PCA,-status_group), scale = T, center = T)
summary(pr_chara)
```

```{r}
screeplot(pr_chara, type = "lines")
```

```{r}
pr_chara$rotation
```

```{r}
ggbiplot(pr_chara, ellipse = TRUE, groups=train$status_group) + coord_equal(ratio=0.35)
```

```{r}
pca_chara_df <- rownames_to_column(as.data.frame(pr_chara$rotation)) %>%
  select(1:5) %>%
  filter(abs(PC1) >= 0.25 | abs(PC2) >= 0.25 | abs(PC3) >= 0.25 | abs(PC4) >= 0.25)

pca_chara_df
```

```{r}
save(pca_chara_df,file="pca_chara_df.Rda")
load("pca_chara_df.Rda")
```

```{r}
# pc1_chara is handpump extraction method, poor water quality, and shallow groundwater wells
# pc2_chara is other extraction method (not handpump, motorpump, and/or gravity), poor or unknown water quality, and surface or river/lake sourced wells 
```

# PC 1

```{r}
prc_chara <- bind_cols(select(train,status_group),as.data.frame(pr_chara$x))

prc_chara %>% select(status_group, PC1) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 2

```{r}
prc_chara %>% select(status_group, PC2) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 3

```{r}
prc_chara %>% select(status_group, PC3) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 4

```{r}
prc_chara %>% select(status_group, PC4) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 5

```{r}
prc_chara %>% select(status_group, PC5) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# PC 6

```{r}
prc_chara %>% select(status_group, PC6) %>%
  pivot_longer(cols = -status_group,names_to = "component",values_to = "loading") %>%
  ggplot(aes(loading, fill = as.factor(status_group))) + geom_density(alpha=0.5)
```

# Selecting features to be used in modeling phase

```{r}
final_train <- train %>% select(status_group, amount_tsh, permit, literacy_rate_by_region, pct_unemployed_by_region, well_strain)

final_train <- final_train %>% dummy_cols(., select_columns = c('permit'), remove_first_dummy = T, remove_selected_columns = T)

final_prc_geo <- prc_geo %>% select(status_group, PC1, PC2) %>% transmute(status_group = status_group, med_lg_SE_region = PC1, small_med_NW_regions = PC2)

final_prc_demo <- prc_demo %>% select(status_group, PC1, PC2) %>% transmute(status_group = status_group, dense_populated_wells = PC1, sparsely_pop_wells = PC2)

final_prc_chara <- prc_chara %>% select(status_group, PC1, PC2) %>% transmute(status_group = status_group, handpump_groundwater_shallow_wells = PC1, other_extraction_from_rivers_lakes = PC2)
```

```{r}
rm(characteristics_PCA, demographic_PCA, geographic_PCA, PCA_variables, pr_chara, pr_demo, pr_geo, pr_train, prc, prc_chara, prc_demo, prc_geo, train)
```

```{r}
final_train <- rowid_to_column(final_train, 'X')
final_prc_chara <- rowid_to_column(final_prc_chara, 'X')
final_prc_demo <- rowid_to_column(final_prc_demo, 'X')
final_prc_geo <- rowid_to_column(final_prc_geo, 'X')
```

```{r}
final_train <- final_train %>% 
  left_join(., final_prc_chara, by = 'X') %>% 
  left_join(., final_prc_demo, by = 'X') %>%
  left_join(., final_prc_geo, by = 'X') %>%
  select(-X, -status_group.y, -status_group.x.x, -status_group.y.y) %>%
  mutate(status_group = status_group.x) %>% 
  select(-status_group.x)
```

```{r}
write.csv(final_train, 'final_train.csv')
```

```{r}
rm(final_prc_chara, final_prc_demo, final_prc_geo)
```

